Learning growth mechanisms of tail realistic preferential attachment models from network degree distributions

### 2025-09-30 Files/paper.qmd
Finalising edits

The popularity of using the power law for network degrees can be traced back to the preferential attachment (PA) model popularised by @Barabasi99. In the general model, as new nodes join the network, an existing node with degree $k$ gains edges at a rate proportional to $b(k)$, where $b(\cdot)$ is a non-negative preference function. @Barabasi99 showed that, when $b(k) = k + 1$, in the limit the resulting degree distribution is regularly varying with index 2. Subsequently, if a real network is shown to be scale-free, one can loosely justify PA as the underlying mechanism of its growth.

@@

This paper aims to address the gap between the applied and theoretical works, by asking if a network is assumed to come from a PA model, can we use the degree distribution alone to directly infer the parameters of the the preference function and learn about the growth mechanisms? Moreover, proper consideration is given to the tail of the degree distribution, because otherwise the effects of the largest degrees, which correspond to the most influential nodes, deviating from the power law will be discounted.

While Voitalov et al. (2019) have pioneered using methods in extreme value theory to analyse degree distributions, they overlooked the inapplicability of standard tools for continuous extremes, due to the discrete nature of degrees. A key difficulty arises because many standard discrete distributions, such as the Poisson, geometric, or negative binomial, do not satisfy the conditions required to belong to a maximum domain of attraction (MDA). Shimura (2012) showed that for a discrete distribution $F$ must be long-tailed in order to be in an MDA, and introduced the notion of recoverability to the MDA, also referred to as the discrete MDA (Hitz et al., 2024). Using the theoretical guarantees of regular variation for a discrete distribution by Shimura (2012), we demonstrate how the tail of the degree distribution is directly influenced by $b(\cdot)$, and subsequently proposed a class of preference functions that not only imply regularly varying degree distributions but are also tail realistic for real networks. These analytical results enable the likelihood of the degree distribution to be expressed in terms of the parameters of $b(\cdot)$, which in turn allows the underlying mechanism of the network, assumed to grow according to PA, to be inferred directly.

The remainder of the paper is as follows: @sec-tail gives a detailed description of the PA model alongside the theoretical results for the survival function of the limiting degree distribution, with a focus on the tail behaviour in terms of the preference function $b(\cdot)$. A class of asymptotically linear preference functions will be introduced and shown to guarantee regular variation in the degree distribution while remaining flexible up until a threshold. @sec-model utilises the proposed preference function and illustrates numerically how the tail index of the degree distribution varies with the model parameters. The simulation study in Subsection -@sec-sim demonstrates that the parameters can be recovered from fitting the model to only the degree distribution. Subsection -@sec-real fits the model to some real data and provides posterior estimates for the preference function. @sec-conc provides a discussion of this paper and possible avenues for future work.

# Tail Behaviour of Preferential Attachment Model {#sec-tail}

@@

Starting at time $t=0$ with an initial network of $m$ vertices that each have no edges, at times $t=1,2,\ldots$ a new vertex is added to the network bringing with it $m$ directed edges from the new vertex; the target for each of these edges are selected from the vertices already in the network with weights proportional to some non-decreasing preference function $b(\cdot)$ of their degree, where $b: \mathbb N \mapsto \mathbb R^+\setminus\{0\}$ is such that:

@@

```{=tex}
Special cases of this model include the Barab\'asi-Albert (BA) model when $b(k) = k+1$, which in the limit of $t\rightarrow \infty$ leads to a power-law degree distribution with tail index 2, and the Uniform Attachment (UA) model where $b(k)=c$ leading to a degree distribution that is not regularly varying.
```
The survival function of the limiting degree distribution, called the limiting survival hereafter, under condition [-@eq-condb2] can be analytically derived in the case where $m=1$, which is presented below.

@@

Now, let $\Upsilon(t)$ be the tree determined by $\zeta(t)$ as follows: $\Upsilon(t)=\{\emptyset\}$ and $\Upsilon(t)=G$ where each existing node $x$ in $\Upsilon(t)$ gives birth to a child with rate $b(\mathrm{deg}(x, \Upsilon(t)))$ independently of the other nodes where $\mathrm{deg}(x, \Upsilon(t))$ is the degree of node $x$ in the tree $\Upsilon(t)$ at time $t$, denote by $\Upsilon(t)_{\downarrow x}$ the tree when treating node $x$ as the root.

@@

$$
\lim_{t\rightarrow\infty}\frac{1}{|\Upsilon(t)|}\sum_{x\in\Upsilon(t)}\varphi(\Upsilon(t)_{\downarrow x}) = \lambda \int_0^\infty e^{-\lambda t}\mathbb E\left[\varphi(\Upsilon(t))\right]\text{d}t.
$$ {#eq-survlim}

@@

@shimura12 established that if $\lim_{k\rightarrow\infty} \Omega(F,k) = 1/\alpha$ ($\alpha>0$), then $F$ is regularly varying with $\bar F(k) \sim k^{-\alpha}$, and hence belongs to the Frechet MDA. If instead, $\lim_{k\rightarrow\infty} \Omega(F,k) = 0$ then the distribution is recoverable to the Gumbel MDA; in this cases we say that the distribution is light-tailed. Moreover, if the distribution is also long-tailed, we can conclude that $F$ itself lies in the Gumbel MDA.

@@

Here, $\lambda^*$ is the Malthusian parameter of the corresponding branching process.See Appendix [-@sec-proofs] for the details of the proof.
:::

@@

@cor-omega2 is a direct consequence of @prp-omega and the results of @shimura12, and it aligns with previous findings in the literature. In particular it is consistent with @krapivsky01, who showed that a sub-linear preference function leads to a light-tailed distributions, since for $b(k)=k^\alpha$ with $\alpha<1$, we have $\lim_{k\rightarrow\infty}[b(k+1)-b(k)]=0$. Conversely, it also agrees with the limiting behaviour of BA model, which produces a regularly degree distribution with tail index 2; for the preference function $b(k)=k+\varepsilon$, $\lim_{k\rightarrow\infty}[b(k+1)-b(k)]=1$, giving a tail index of $\lambda^*=2$

@@

which has to be solved numerically for most parameter choices. Also, note that $\lambda>\beta$.

@@

```{r, echo=FALSE, warning=FALSE, cache=TRUE, out.width="119mm"}
#| fig-width: 8
#| fig-height: 4
#| fig-cap: The limiting survival, according to various combinations of $(\alpha, \beta, \varepsilon)$ and $k_0=20$ of the proposed PA model.
#| label: fig-polylinsurv

@@

\begin{align*}
L(\pmb n | \pmb \theta,l) = &\left(\frac{\lambda^*}{\lambda^*+\varepsilon}\right)^{n_0}\left(\prod_{j=l}^{k_0-1}\frac{j^\alpha +\varepsilon}{\lambda^* + j^\alpha +\varepsilon}\right)^{\left(\sum_{i\ge k_0}n_{i}\right)} \\ &\times \prod_{l \le i<k_0}\left(\frac{\lambda^*}{\lambda^* +i^\alpha + \varepsilon } \prod_{j=l}^{k_0-1}\frac{j^\alpha + \varepsilon}{\lambda^* + j^\alpha + \varepsilon}\right)^{n_i}\\ &\times \prod_{i\ge k_0}\left(\frac{\text{B}(i-k_0 + (k_0^\alpha + \varepsilon)/\beta,1+\lambda^*/\beta)}{\text{B}((k_0^\alpha + \varepsilon)/\beta,\lambda^*/\beta)}\right)^{n_i}
\end{align*}\label{eq-lh}

where $\text{B}(\cdot,\cdot)$ is the the beta function, $\pmb \theta = (\alpha, \varepsilon, k_0,\beta)$, and $l\ge0$ is a quantity that allows truncating the data such that the minimum degree is $l$. This will allow the model to be fitted whilst ignoring the influence of the lower degrees (those less than $l$) as the model does not capture the behaviour well at the lower degrees, since @rudas07 only provides results for the case of a preferential attachment tree.

@@

This subsection aims to show that the parameters of the model (and therefore the preference function) in @sec-model can be recovered from simulating a network from the model, and fitting it to the observed degree distribution, using the likelihood in \eqref{eq-lh}.

@@

where Gamma($a$, $b$) is the gamma distribution with shape $a$ and rate $b$, and U($a$, $b$) is uniform distribution with lower and upper bounds $a$ and $b$, to obtain a posterior distribution, up to the proportionality constant. Posterior samples can then be obtained by an adaptive Metropolis-Hastings Markov chain Monte Carlo (MCMC) algorithm. For these simulated networks $l=0$.

@@

\begin{align*}
\sum_{n=0}^\infty\frac{\Gamma(n+x)}{\Gamma(n+x+y)} &=\frac{1}{\Gamma(y)}\sum_{n=0}^\infty \text{B}(n+x,y)\\
&=\frac{1}{\Gamma(y)}\sum_{n=0}^\infty\int_0^1t^{n+x-1}(1-t)^{y-1}at\\
&=\frac{1}{\Gamma(y)}\int_0^1 t^{x-1}(1-t)^{y-1}\sum_{n=0}^\infty t^n\qquad at\\
&=\frac{1}{\Gamma(y)}\int_0^1 t^{x-1}(1-t)^{y-1}\frac{1}{1-t}at\\
&=\frac{1}{\Gamma(y)}\int_0^1 t^{x-1}(1-t)^{y-2}at\\
&=\frac{1}{\Gamma(y)}\text{y}(x,y-1)\\
&= \frac{\Gamma(x)}{(y-1)\Gamma(x+y-1)}.
\end{align*}

>>>>

```{=tex}
The popularity of using the power law for network degrees can be traced back to a preferential attachment (PA) model popularised by \cite{Barabasi99}. In the general model, general preferential attachment (GPA), as new nodes join the network, an existing node with degree $k$ gains edges at a rate proportional to $b(k)$, where $b(\cdot)$ is a non-negative preference function. \cite{Barabasi99} showed that, when $b(k) = k + 1$, in the limit the resulting degree distribution is regularly varying with index 2 — this model is commonly known as the Barab\'asi-Albert (BA) model.  Subsequently, if a real network is shown to be scale-free, one can loosely justify PA as the underlying mechanism of its growth.
```

@@

This paper aims to address the gap between the applied and theoretical works, by asking if a network is assumed to come from a GPA model, can we use the degree distribution alone to directly infer the parameters of the the preference function and learn about the growth mechanisms? Moreover, proper consideration is given to the tail of the degree distribution, because otherwise the effects of the largest degrees, which correspond to the most influential nodes, deviating from the power law will be discounted.

While @Voitalov_2019 have pioneered using methods in extreme value theory to analyse degree distributions, they overlooked the inapplicability of standard tools for continuous extremes, due to the discrete nature of degrees. A key difficulty arises because many standard discrete distributions, such as the Poisson, geometric, or negative binomial, do not satisfy the conditions required to belong to a maximum domain of attraction (MDA). @shimura12 showed that a discrete distribution $F$ must be long-tailed in order to be in an MDA, and introduced the notion of recoverability to the MDA, also referred to as the discrete MDA [@hitz24]. Using the theoretical guarantees of regular variation for a discrete distribution by @shimura12, we demonstrate how the tail of the degree distribution is directly influenced by $b(\cdot)$, and subsequently propose a class of preference functions that not only imply regularly varying degree distributions but are also tail realistic for real networks. These analytical results enable the likelihood of the degree distribution to be expressed in terms of the parameters of $b(\cdot)$, which in turn allows the underlying mechanism of the network, assumed to grow according to GPA, to be inferred directly.

The remainder of the paper is as follows: @sec-tail gives a detailed description of the GPA model alongside the theoretical results for the survival function of the limiting degree distribution, with a focus on the tail behaviour in terms of the preference function $b(\cdot)$. A class of asymptotically linear preference functions will be introduced and shown to guarantee regular variation in the degree distribution while remaining flexible up until a threshold. @sec-model utilises the proposed preference function and illustrates numerically how the tail index of the degree distribution varies with the model parameters. The simulation study in @sec-sim demonstrates that the parameters can be recovered from fitting the model to only the degree distribution. @sec-real fits the model to some real data and provides posterior estimates for the preference function. @sec-conc provides a discussion of this paper and possible avenues for future work.

# Tail Behaviour of GPA Model {#sec-tail}

@@

Starting at time $t=0$ with an initial network of $m$ vertices that each have no edges, at times $t=1,2,\ldots$ a new vertex is added to the network bringing with it $m$ directed edges from the new vertex; the targets for each of these edges are selected from the vertices already in the network with weights proportional to some non-decreasing preference function $b(\cdot)$ of their degree, where $b: \mathbb N \mapsto \mathbb R^+\setminus\{0\}$ is such that:

@@

```{=tex}
Special cases of this model include the BA model when $b(k) = k+1$, which in the limit of $t\rightarrow \infty$ leads to a power-law degree distribution with tail index 2, and the Uniform Attachment (UA) model where $b(k)=c$ leading to a degree distribution that is not regularly varying.
```
The survival function of the limiting degree distribution, called the limiting survival hereafter, under condition [-@eq-condb2] can be analytically derived in the case where $m=1$, which is presented below.

@@

Now, let $\Upsilon(t)$ be the tree determined by $\zeta(t)$ as follows: $\Upsilon(0)=\{\emptyset\}$ and $\Upsilon(t)$ is a graph where each existing node $x$ in $\Upsilon(t)$ gives birth to a child with rate $b(\mathrm{deg}(x, \Upsilon(t)))$ independently of the other nodes where $\mathrm{deg}(x, \Upsilon(t))$ is the degree of node $x$ in the tree $\Upsilon(t)$ at time $t$, denote by $\Upsilon(t)_{\downarrow x}$ the tree when treating node $x$ as the root.

@@

$$
\lim_{t\rightarrow\infty}\frac{1}{|\Upsilon(t)|}\sum_{x\in\Upsilon(t)}\varphi(\Upsilon(t)_{\downarrow x}) = \lambda^* \int_0^\infty e^{-\lambda^* t}\mathbb E\left[\varphi(\Upsilon(t))\right]\text{d}t.
$$ {#eq-survlim}

@@

@shimura12 established that if $\lim_{k\rightarrow\infty} \Omega(F,k) = 1/\alpha$ ($\alpha>0$), then $F$ is regularly varying with $\bar F(k) \sim k^{-\alpha}$, and hence belongs to the Fréchet MDA. If instead, $\lim_{k\rightarrow\infty} \Omega(F,k) = 0$ then the distribution is recoverable to the Gumbel MDA; in this cases we say that the distribution is light-tailed. Moreover, if the distribution is also long-tailed, we can conclude that $F$ itself lies in the Gumbel MDA.

@@

Here, $\lambda^*$ is the Malthusian parameter of the corresponding branching process. See Appendix [-@sec-proofs] for the details of the proof.
:::

@@

@cor-omega2 is a direct consequence of @prp-omega and the results of @shimura12, and it aligns with previous findings in the literature. In particular it is consistent with @krapivsky01, who showed that a sub-linear preference function leads to a light-tailed distributions, since for $b(k)=k^\alpha$ with $\alpha<1$, we have $\lim_{k\rightarrow\infty}[b(k+1)-b(k)]=0$. Conversely, it also agrees with the limiting behaviour of the BA model, which produces a regularly degree distribution with tail index 2; for the preference function $b(k)=k+1$, $\lim_{k\rightarrow\infty}[b(k+1)-b(k)]=1$, giving a tail index of $\lambda^*=2$.

@@

which has to be solved numerically for most parameter choices. Also, note that $\lambda>\beta$. See Appendix [-@sec-proofs] for the derivations of @eq-rho.

@@

```{r, echo=FALSE, warning=FALSE, cache=TRUE, out.width="119mm"}
#| fig-width: 8
#| fig-height: 4
#| fig-cap: The limiting survival, according to various combinations of $(\alpha, \beta, \varepsilon)$ and $k_0=20$ of the proposed GPA model.
#| label: fig-polylinsurv

@@

$$
\begin{aligned}
L(\pmb n | \pmb \theta,l) = &\left(\frac{\lambda^*}{\lambda^*+\varepsilon}\right)^{n_0}\left(\prod_{j=l}^{k_0-1}\frac{j^\alpha +\varepsilon}{\lambda^* + j^\alpha +\varepsilon}\right)^{\left(\sum_{i\ge k_0}n_{i}\right)} \\ &\times \prod_{l \le i<k_0}\left(\frac{\lambda^*}{\lambda^* +i^\alpha + \varepsilon } \prod_{j=l}^{k_0-1}\frac{j^\alpha + \varepsilon}{\lambda^* + j^\alpha + \varepsilon}\right)^{n_i}\\ &\times \prod_{i\ge k_0}\left(\frac{\text{B}(i-k_0 + (k_0^\alpha + \varepsilon)/\beta,1+\lambda^*/\beta)}{\text{B}((k_0^\alpha + \varepsilon)/\beta,\lambda^*/\beta)}\right)^{n_i}
\end{aligned}
$${#eq-llh}
where $\text{B}(\cdot,\cdot)$ is the the beta function, $\pmb \theta = (\alpha, \varepsilon, k_0,\beta)$, and $l\ge0$ is a quantity that allows truncating the data such that the minimum degree is $l$. This will allow the model to be fitted whilst ignoring the influence of the lower degrees (those less than $l$) as the model does not capture the behaviour well at the lower degrees, since @rudas07 only provides results for the case of a preferential attachment tree.

@@

This subsection aims to show that the parameters of the proposed model (and therefore the preference function) in @sec-model can be recovered from simulating a network from the model, and fitting it to the observed degree distribution, using the likelihood in @eq-llh.

@@

where Gamma($a$, $b$) is the gamma distribution with shape $a$ and rate $b$, and U($a$, $b$) is discrete uniform distribution with lower and upper bounds $a$ and $b$, to obtain a posterior distribution, up to the proportionality constant. Posterior samples can then be obtained by an adaptive Metropolis-Hastings Markov chain Monte Carlo (MCMC) algorithm. For these simulated networks $l=0$.

@@

\begin{align*}
\sum_{n=0}^\infty\frac{\Gamma(n+x)}{\Gamma(n+x+y)} &=\frac{1}{\Gamma(y)}\sum_{n=0}^\infty \text{B}(n+x,y)\\
&=\frac{1}{\Gamma(y)}\sum_{n=0}^\infty\int_0^1t^{n+x-1}(1-t)^{y-1}\text{d}t\\
&=\frac{1}{\Gamma(y)}\int_0^1 t^{x-1}(1-t)^{y-1}\sum_{n=0}^\infty t^n \text{d}t\\
&=\frac{1}{\Gamma(y)}\int_0^1 t^{x-1}(1-t)^{y-1}\frac{1}{1-t}\text{d}t\\
&=\frac{1}{\Gamma(y)}\int_0^1 t^{x-1}(1-t)^{y-2}at\\
&=\frac{1}{\Gamma(y)}\text{B}(x,y-1)\\
&= \frac{\Gamma(x)}{(y-1)\Gamma(x+y-1)}.
\end{align*}

### 2025-09-30 Files/paper.tex
Finalising edits

The popularity of using the power law for network degrees can be traced
back to the preferential attachment (PA) model popularised by
\citet{Barabasi99}. In the general model, as new nodes join the network,
an existing node with degree \(k\) gains edges at a rate proportional to
\(b(k)\), where \(b(\cdot)\) is a non-negative preference function.
\citet{Barabasi99} showed that, when \(b(k) = k + 1\), in the limit the
resulting degree distribution is regularly varying with index 2.
Subsequently, if a real network is shown to be scale-free, one can
loosely justify PA as the underlying mechanism of its growth.

@@

This paper aims to address the gap between the applied and theoretical
works, by asking if a network is assumed to come from a PA model, can we
use the degree distribution alone to directly infer the parameters of
the the preference function and learn about the growth mechanisms?
Moreover, proper consideration is given to the tail of the degree
distribution, because otherwise the effects of the largest degrees,
which correspond to the most influential nodes, deviating from the power
law will be discounted.

While Voitalov et al.~(2019) have pioneered using methods in extreme
value theory to analyse degree distributions, they overlooked the
inapplicability of standard tools for continuous extremes, due to the
discrete nature of degrees. A key difficulty arises because many
standard discrete distributions, such as the Poisson, geometric, or
negative binomial, do not satisfy the conditions required to belong to a
maximum domain of attraction (MDA). Shimura (2012) showed that for a
discrete distribution \(F\) must be long-tailed in order to be in an
MDA, and introduced the notion of recoverability to the MDA, also
referred to as the discrete MDA (Hitz et al., 2024). Using the
theoretical guarantees of regular variation for a discrete distribution
by Shimura (2012), we demonstrate how the tail of the degree
distribution is directly influenced by \(b(\cdot)\), and subsequently
proposed a class of preference functions that not only imply regularly
varying degree distributions but are also tail realistic for real
networks. These analytical results enable the likelihood of the degree
distribution to be expressed in terms of the parameters of \(b(\cdot)\),
which in turn allows the underlying mechanism of the network, assumed to
grow according to PA, to be inferred directly.

The remainder of the paper is as follows: Section~\ref{sec-tail} gives a
detailed description of the PA model alongside the theoretical results
for the survival function of the limiting degree distribution, with a
focus on the tail behaviour in terms of the preference function
\(b(\cdot)\). A class of asymptotically linear preference functions will
be introduced and shown to guarantee regular variation in the degree
distribution while remaining flexible up until a threshold.
Section~\ref{sec-model} utilises the proposed preference function and
illustrates numerically how the tail index of the degree distribution
varies with the model parameters. The simulation study in Subsection
\ref{sec-sim} demonstrates that the parameters can be recovered from
fitting the model to only the degree distribution. Subsection
\ref{sec-real} fits the model to some real data and provides posterior
estimates for the preference function. Section~\ref{sec-conc} provides a
discussion of this paper and possible avenues for future work.

\section{Tail Behaviour of Preferential Attachment
Model}\label{sec-tail}

@@

Starting at time \(t=0\) with an initial network of \(m\) vertices that
each have no edges, at times \(t=1,2,\ldots\) a new vertex is added to
the network bringing with it \(m\) directed edges from the new vertex;
the target for each of these edges are selected from the vertices
already in the network with weights proportional to some non-decreasing
preference function \(b(\cdot)\) of their degree, where
\(b: \mathbb N \mapsto \mathbb R^+\setminus\{0\}\) is such that:

@@

Special cases of this model include the Barab\'asi-Albert (BA) model when $b(k) = k+1$, which in the limit of $t\rightarrow \infty$ leads to a power-law degree distribution with tail index 2, and the Uniform Attachment (UA) model where $b(k)=c$ leading to a degree distribution that is not regularly varying.

@@

Now, let \(\Upsilon(t)\) be the tree determined by \(\zeta(t)\) as
follows: \(\Upsilon(t)=\{\emptyset\}\) and \(\Upsilon(t)=G\) where each
existing node \(x\) in \(\Upsilon(t)\) gives birth to a child with rate
\(b(\mathrm{deg}(x, \Upsilon(t)))\) independently of the other nodes
where \(\mathrm{deg}(x, \Upsilon(t))\) is the degree of node \(x\) in
the tree \(\Upsilon(t)\) at time \(t\), denote by
\(\Upsilon(t)_{\downarrow x}\) the tree when treating node \(x\) as the
root.

@@

\begin{equation}\phantomsection\label{eq-survlim}{
\lim_{t\rightarrow\infty}\frac{1}{|\Upsilon(t)|}\sum_{x\in\Upsilon(t)}\varphi(\Upsilon(t)_{\downarrow x}) = \lambda \int_0^\infty e^{-\lambda t}\mathbb E\left[\varphi(\Upsilon(t))\right]\text{d}t.
}\end{equation}

@@

\citet{shimura12} established that if
\(\lim_{k\rightarrow\infty} \Omega(F,k) = 1/\alpha\) (\(\alpha>0\)),
then \(F\) is regularly varying with \(\bar F(k) \sim k^{-\alpha}\), and
hence belongs to the Frechet MDA. If instead,
\(\lim_{k\rightarrow\infty} \Omega(F,k) = 0\) then the distribution is
recoverable to the Gumbel MDA; in this cases we say that the
distribution is light-tailed. Moreover, if the distribution is also
long-tailed, we can conclude that \(F\) itself lies in the Gumbel MDA.

@@

Here, \(\lambda^*\) is the Malthusian parameter of the corresponding
branching process.See Appendix \ref{sec-proofs} for the details of the
proof.

@@

Corollary~\ref{cor-omega2} is a direct consequence of
Proposition~\ref{prp-omega} and the results of \citet{shimura12}, and it
aligns with previous findings in the literature. In particular it is
consistent with \citet{krapivsky01}, who showed that a sub-linear
preference function leads to a light-tailed distributions, since for
\(b(k)=k^\alpha\) with \(\alpha<1\), we have
\(\lim_{k\rightarrow\infty}[b(k+1)-b(k)]=0\). Conversely, it also agrees
with the limiting behaviour of BA model, which produces a regularly
degree distribution with tail index 2; for the preference function
\(b(k)=k+\varepsilon\), \(\lim_{k\rightarrow\infty}[b(k+1)-b(k)]=1\),
giving a tail index of \(\lambda^*=2\)

@@

which has to be solved numerically for most parameter choices. Also,
note that \(\lambda>\beta\).

@@

\caption{\label{fig-polylinsurv}The limiting survival, according to
various combinations of \((\alpha, \beta, \varepsilon)\) and \(k_0=20\)
of the proposed PA model.}

@@

\begin{align*}
L(\pmb n | \pmb \theta,l) = &\left(\frac{\lambda^*}{\lambda^*+\varepsilon}\right)^{n_0}\left(\prod_{j=l}^{k_0-1}\frac{j^\alpha +\varepsilon}{\lambda^* + j^\alpha +\varepsilon}\right)^{\left(\sum_{i\ge k_0}n_{i}\right)} \\ &\times \prod_{l \le i<k_0}\left(\frac{\lambda^*}{\lambda^* +i^\alpha + \varepsilon } \prod_{j=l}^{k_0-1}\frac{j^\alpha + \varepsilon}{\lambda^* + j^\alpha + \varepsilon}\right)^{n_i}\\ &\times \prod_{i\ge k_0}\left(\frac{\text{B}(i-k_0 + (k_0^\alpha + \varepsilon)/\beta,1+\lambda^*/\beta)}{\text{B}((k_0^\alpha + \varepsilon)/\beta,\lambda^*/\beta)}\right)^{n_i}
\end{align*}\label{eq-lh}

where \(\text{B}(\cdot,\cdot)\) is the the beta function,
\(\pmb \theta = (\alpha, \varepsilon, k_0,\beta)\), and \(l\ge0\) is a
quantity that allows truncating the data such that the minimum degree is
\(l\). This will allow the model to be fitted whilst ignoring the
influence of the lower degrees (those less than \(l\)) as the model does
not capture the behaviour well at the lower degrees, since
\citet{rudas07} only provides results for the case of a preferential
attachment tree.

@@

This subsection aims to show that the parameters of the model (and
therefore the preference function) in Section~\ref{sec-model} can be
recovered from simulating a network from the model, and fitting it to
the observed degree distribution, using the likelihood in \eqref{eq-lh}.

@@

where Gamma(\(a\), \(b\)) is the gamma distribution with shape \(a\) and
rate \(b\), and U(\(a\), \(b\)) is uniform distribution with lower and
upper bounds \(a\) and \(b\), to obtain a posterior distribution, up to
the proportionality constant. Posterior samples can then be obtained by
an adaptive Metropolis-Hastings Markov chain Monte Carlo (MCMC)
algorithm. For these simulated networks \(l=0\).

>>>>

The popularity of using the power law for network degrees can be traced back to a preferential attachment (PA) model popularised by \cite{Barabasi99}. In the general model, general preferential attachment (GPA), as new nodes join the network, an existing node with degree $k$ gains edges at a rate proportional to $b(k)$, where $b(\cdot)$ is a non-negative preference function. \cite{Barabasi99} showed that, when $b(k) = k + 1$, in the limit the resulting degree distribution is regularly varying with index 2 — this model is commonly known as the Barab\'asi-Albert (BA) model.  Subsequently, if a real network is shown to be scale-free, one can loosely justify PA as the underlying mechanism of its growth.

@@

This paper aims to address the gap between the applied and theoretical
works, by asking if a network is assumed to come from a GPA model, can
we use the degree distribution alone to directly infer the parameters of
the the preference function and learn about the growth mechanisms?
Moreover, proper consideration is given to the tail of the degree
distribution, because otherwise the effects of the largest degrees,
which correspond to the most influential nodes, deviating from the power
law will be discounted.

While \citet{Voitalov_2019} have pioneered using methods in extreme
value theory to analyse degree distributions, they overlooked the
inapplicability of standard tools for continuous extremes, due to the
discrete nature of degrees. A key difficulty arises because many
standard discrete distributions, such as the Poisson, geometric, or
negative binomial, do not satisfy the conditions required to belong to a
maximum domain of attraction (MDA). \citet{shimura12} showed that a
discrete distribution \(F\) must be long-tailed in order to be in an
MDA, and introduced the notion of recoverability to the MDA, also
referred to as the discrete MDA \citep{hitz24}. Using the theoretical
guarantees of regular variation for a discrete distribution by
\citet{shimura12}, we demonstrate how the tail of the degree
distribution is directly influenced by \(b(\cdot)\), and subsequently
propose a class of preference functions that not only imply regularly
varying degree distributions but are also tail realistic for real
networks. These analytical results enable the likelihood of the degree
distribution to be expressed in terms of the parameters of \(b(\cdot)\),
which in turn allows the underlying mechanism of the network, assumed to
grow according to GPA, to be inferred directly.

The remainder of the paper is as follows: Section~\ref{sec-tail} gives a
detailed description of the GPA model alongside the theoretical results
for the survival function of the limiting degree distribution, with a
focus on the tail behaviour in terms of the preference function
\(b(\cdot)\). A class of asymptotically linear preference functions will
be introduced and shown to guarantee regular variation in the degree
distribution while remaining flexible up until a threshold.
Section~\ref{sec-model} utilises the proposed preference function and
illustrates numerically how the tail index of the degree distribution
varies with the model parameters. The simulation study in
Section~\ref{sec-sim} demonstrates that the parameters can be recovered
from fitting the model to only the degree distribution.
Section~\ref{sec-real} fits the model to some real data and provides
posterior estimates for the preference function. Section~\ref{sec-conc}
provides a discussion of this paper and possible avenues for future
work.

\section{Tail Behaviour of GPA Model}\label{sec-tail}

@@

Starting at time \(t=0\) with an initial network of \(m\) vertices that
each have no edges, at times \(t=1,2,\ldots\) a new vertex is added to
the network bringing with it \(m\) directed edges from the new vertex;
the targets for each of these edges are selected from the vertices
already in the network with weights proportional to some non-decreasing
preference function \(b(\cdot)\) of their degree, where
\(b: \mathbb N \mapsto \mathbb R^+\setminus\{0\}\) is such that:

@@

Special cases of this model include the BA model when $b(k) = k+1$, which in the limit of $t\rightarrow \infty$ leads to a power-law degree distribution with tail index 2, and the Uniform Attachment (UA) model where $b(k)=c$ leading to a degree distribution that is not regularly varying.

@@

Now, let \(\Upsilon(t)\) be the tree determined by \(\zeta(t)\) as
follows: \(\Upsilon(0)=\{\emptyset\}\) and \(\Upsilon(t)\) is a graph
where each existing node \(x\) in \(\Upsilon(t)\) gives birth to a child
with rate \(b(\mathrm{deg}(x, \Upsilon(t)))\) independently of the other
nodes where \(\mathrm{deg}(x, \Upsilon(t))\) is the degree of node \(x\)
in the tree \(\Upsilon(t)\) at time \(t\), denote by
\(\Upsilon(t)_{\downarrow x}\) the tree when treating node \(x\) as the
root.

@@

\begin{equation}\phantomsection\label{eq-survlim}{
\lim_{t\rightarrow\infty}\frac{1}{|\Upsilon(t)|}\sum_{x\in\Upsilon(t)}\varphi(\Upsilon(t)_{\downarrow x}) = \lambda^* \int_0^\infty e^{-\lambda^* t}\mathbb E\left[\varphi(\Upsilon(t))\right]\text{d}t.
}\end{equation}

@@

\citet{shimura12} established that if
\(\lim_{k\rightarrow\infty} \Omega(F,k) = 1/\alpha\) (\(\alpha>0\)),
then \(F\) is regularly varying with \(\bar F(k) \sim k^{-\alpha}\), and
hence belongs to the Fréchet MDA. If instead,
\(\lim_{k\rightarrow\infty} \Omega(F,k) = 0\) then the distribution is
recoverable to the Gumbel MDA; in this cases we say that the
distribution is light-tailed. Moreover, if the distribution is also
long-tailed, we can conclude that \(F\) itself lies in the Gumbel MDA.

@@

Here, \(\lambda^*\) is the Malthusian parameter of the corresponding
branching process. See Appendix \ref{sec-proofs} for the details of the
proof.

@@

Corollary~\ref{cor-omega2} is a direct consequence of
Proposition~\ref{prp-omega} and the results of \citet{shimura12}, and it
aligns with previous findings in the literature. In particular it is
consistent with \citet{krapivsky01}, who showed that a sub-linear
preference function leads to a light-tailed distributions, since for
\(b(k)=k^\alpha\) with \(\alpha<1\), we have
\(\lim_{k\rightarrow\infty}[b(k+1)-b(k)]=0\). Conversely, it also agrees
with the limiting behaviour of the BA model, which produces a regularly
degree distribution with tail index 2; for the preference function
\(b(k)=k+1\), \(\lim_{k\rightarrow\infty}[b(k+1)-b(k)]=1\), giving a
tail index of \(\lambda^*=2\).

@@

which has to be solved numerically for most parameter choices. Also,
note that \(\lambda>\beta\). See Appendix \ref{sec-proofs} for the
derivations of Equation~\ref{eq-rho}.

@@

\caption{\label{fig-polylinsurv}The limiting survival, according to
various combinations of \((\alpha, \beta, \varepsilon)\) and \(k_0=20\)
of the proposed GPA model.}

@@

\begin{equation}\phantomsection\label{eq-llh}{
\begin{aligned}
L(\pmb n | \pmb \theta,l) = &\left(\frac{\lambda^*}{\lambda^*+\varepsilon}\right)^{n_0}\left(\prod_{j=l}^{k_0-1}\frac{j^\alpha +\varepsilon}{\lambda^* + j^\alpha +\varepsilon}\right)^{\left(\sum_{i\ge k_0}n_{i}\right)} \\ &\times \prod_{l \le i<k_0}\left(\frac{\lambda^*}{\lambda^* +i^\alpha + \varepsilon } \prod_{j=l}^{k_0-1}\frac{j^\alpha + \varepsilon}{\lambda^* + j^\alpha + \varepsilon}\right)^{n_i}\\ &\times \prod_{i\ge k_0}\left(\frac{\text{B}(i-k_0 + (k_0^\alpha + \varepsilon)/\beta,1+\lambda^*/\beta)}{\text{B}((k_0^\alpha + \varepsilon)/\beta,\lambda^*/\beta)}\right)^{n_i}
\end{aligned}
}\end{equation} where \(\text{B}(\cdot,\cdot)\) is the the beta
function, \(\pmb \theta = (\alpha, \varepsilon, k_0,\beta)\), and
\(l\ge0\) is a quantity that allows truncating the data such that the
minimum degree is \(l\). This will allow the model to be fitted whilst
ignoring the influence of the lower degrees (those less than \(l\)) as
the model does not capture the behaviour well at the lower degrees,
since \citet{rudas07} only provides results for the case of a
preferential attachment tree.

@@

This subsection aims to show that the parameters of the proposed model
(and therefore the preference function) in Section~\ref{sec-model} can
be recovered from simulating a network from the model, and fitting it to
the observed degree distribution, using the likelihood in
Equation~\ref{eq-llh}.

@@

where Gamma(\(a\), \(b\)) is the gamma distribution with shape \(a\) and
rate \(b\), and U(\(a\), \(b\)) is discrete uniform distribution with
lower and upper bounds \(a\) and \(b\), to obtain a posterior
distribution, up to the proportionality constant. Posterior samples can
then be obtained by an adaptive Metropolis-Hastings Markov chain Monte
Carlo (MCMC) algorithm. For these simulated networks \(l=0\).
